---
title: "R Tips & Tricks"
author: "Julia Ponomarenko, CRG Bioinformatics core."
date: "Dec 27, 2017"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    theme: readable
---
<br/>

* Nice R tutorial
http://journals.plos.org/ploscompbiol/article/file?id=10.1371/journal.pcbi.1000482.s001&type=supplementary
* R swirl tutorials https://github.com/swirldev/swirl_courses. Basic one was used here.
<br/>

The example dataset used in this tutorial (Davis from the package car)
------------------
```{r, warning=FALSE, message=FALSE}
#install.packages("car")
library(car)
data <- Davis
```

<br/>

Files and Directories
--------------------------
```{r}
getwd() # current directory - This is directory where this file is located
# to set up this directory as a working in console
dir <- getwd()
setwd(dir) # run it in console, not from inside the notebook

file.info("R_Tips.Rmd")
file.info("R_Tips.Rmd")$ctime 
list.files()

dir.create("testdir", showWarnings = F)
setwd("testdir") # run in console
file.create("mytest.R")
file.exists("mytest.R")
file.rename("mytest.R","mytest2.R")
file.copy("mytest2.R","mytest3.R")
list.files()

# get full paths of all files and directories containing "R" or "r" in the directory dir
filenames <- list.files(dir, pattern = "*R*", full.names = TRUE, recursive = TRUE, ignore.case = TRUE, include.dirs = TRUE) 
Dir_separator <- "/" # for Windows it is "\"
file_name <- as.character(lapply(strsplit(filenames[1], Dir_separator), tail, n = 1)) # to get just the file name, for example
# while full.names = FALSE in list.files()does the same without the need to parse filenames

list.dirs() # also useful command when looping via directories is needed


dir.create(file.path('testdir2', 'testdir3'), recursive = TRUE)
setwd("..") # one directory up
unlink("testdir", recursive = T)  #removes testdir and all folders inside
```
<br/>


Read files
-----------
```{r}
file <- "R_Tips.Rmd"
x <- read.table(file, skip = 7, nrows = 1) # for example, we read only line 8
x <- as.numeric(x[2]) # for example, we need a value of the 2nd element of the vector x

# let's write data to the file with the description lines on top
file <- "Davis"
conn <- file(file, open = "w")
writeLines("This is the dataset Davis from the package car", conn)
write("This is my favorite dataset", conn, append = TRUE)
close(conn)

conn <- file(file, open = "a")
write.table(data, conn, row.names = FALSE) # can be write.csv()
close(conn)

#to read a data frame 
conn <- file(file, open = "r")
df <- read.table(conn, skip = 2, header = TRUE) # can be read.csv()
close(conn)


```
<br/>


Data structures: Sequences, Numbers, Vectors, Matrices, Lists  
-----------
```{r}
w <<- 10 # assign w to a global environment ???

6.03 / 2
6.03 %/% 2 # integer division
6.03 %% 2 # getting remainder

my_seq <- seq(5,10,length=30)
my_seq

# all three below are identical
1:length(my_seq)
seq(along.with = my_seq)
seq_along(my_seq)

# notice difference!
rep(c(1,2,3), times = 10)
rep(c(1,2,3), each = 10)

x <- c(1,2,3,4,5,6)
y <- x > 3 # boolean vector

chars <- c("my", "name", "is")
x <- paste(chars, collapse = " ") # concatenate vector
paste("my", "name", "is", sep = " ") # same as above, but not paste(chars, sep = " ")

# Careful: paste with vectors recycles the shorter vector
paste(LETTERS, 1:4, sep = "-") # output is a vector!

# random sampling from vectors without replacement
y <- rnorm(1000)
z <- rep(NA, 1000)
my_data <- sample(c(y, z), 100)
my_na <- is.na(my_data)
sum(my_na) # how many NAs
x <- my_data
y <- x[!is.na(x)]
y[y > 0]
x[!is.na(x) & x > 0] # same as above
x[x > 0] # gives NA for NA because NA > 0 = NA

# NaN stand for "not a number"
# Inf stand for infinity

my_vector <- 1:20
dim(my_vector) <- c(4, 5) # dim allows both to get and set the 'dim' attribute to make a vector to a matrix
attributes(my_vector)
class(my_vector)
my_matrix <- my_vector

my_matrix2 <- matrix(data = 1:20, nrow = 4)
identical(my_matrix, my_matrix2)

# naming matrix rows and columns
patients <- c("Bill", "Gina", "Kelly", "Sean") # are in rows
cbind(patients,my_matrix) # numbers in my_matrix are implicitly coerced in characters!
my_data <- data.frame(patients, my_matrix) # so, use a data.frame instead
cnames <- c("patient", "age", "weight", "bp", "rating", "test")
colnames(my_data) <- cnames

# which() function finds the indices:
ints <- sample(10)
which(ints > 7)
any(ints < 0)
all(ints > 0)
```

#### Lists
```{r}
l <- list(id = "Joe", age = 25, dob = c(1980, 3, 29))
l$id
l[[1]]
l["id"] <- "Jim" #change element
l["age"] <- NULL # drop element
```

#### Operators &, &&, | and ||
```{r}
#the `&` operator to evaluate AND across a vector:
TRUE & c(TRUE, FALSE, FALSE)
#The `&&` version of AND only evaluates the first member of a vector:
TRUE && c(TRUE, FALSE, FALSE)

#The OR operator follows a similar set of rules. The `|` version of OR evaluates OR across an entire vector, while the `||` version of OR only evaluates the first member of a vector.
TRUE | c(TRUE, FALSE, FALSE)

#All AND operators are evaluated before OR operators

isTRUE(6 > 4)
xor(5 == 6, !FALSE)

```
<br/>


Functions
-----------
Type the function name to see its source code.
Use args(function_name) to see the function arguments
<br/>


#### Function can be passed as an argument to other functions
```{r}
evaluate <- function(func, dat){
  return(func(dat))
}
evaluate(sd, c(1.4, 3.6, 7.9, 8.8))
evaluate(function(x){x+1}, 6)
evaluate(function(x){x[1]}, c(8, 4, 0))
evaluate(function(x){x[length(x)]}, c(8, 4, 0)) # accessing the last element
```
<br/>


#### Ellipses
Use ellipsis when the function calls the variable number of arguments, see https://www.r-bloggers.com/r-three-dots-ellipsis/

The ellipses can be used to pass on arguments to other functions that are
used within the function. Usually a function that has the
ellipses as an argument has the ellipses as the last argument. The usage of
such a function would look like
__ellipses_func(arg1, arg2 = TRUE, ...)__
<br/>
However, the usage for the paste() function is as follows:
__paste(..., sep = " ", collapse = NULL)__,
where the ellipses is the first argument, and all other arguments after the ellipses have default values. __This is a strict rule in R programming: all arguments after an ellipses must have default values. __
```{r}
telegram <- function(...){
  args <- list(...)
  txt <- paste(args, collapse = " ")
  paste("START", txt, "STOP", collapse = " ")
}
telegram("I", "want", "to eat")


telegram2 <- function(...){
  args <- list(...)
  #extract named arguments from the args list by the name of the argument and double brackets: 
  who <- args[["who"]]
  verb <- args[["verb"]]
  what <- args[["what"]]
  paste("The person", who, "decided that she", verb, "to do the following", what)
}
telegram2(who = "I", verb = "want", what = "to eat")

```

<br/>


Vectorization - Looping through an entire vector
--------------------------
```{r}
x <- runif(100) # a vector of uniformly distributed 100 values
y <- ifelse(x > 0.5, 1, 0) 
```
<br/>


Q. Flip a biased coin [p=0.6 of heads] 100 times; how many heads do you get? Repeat this for 1000 trials.
```{r}
x <- replicate(1000, rbinom(n = 100, size = 1, p = 0.6))
heads <- colSums(x)
hist(heads)

# or the same
hist(rbinom(n = 1000, size = 100, p = 0.6))
```
<br/>


Q. Fibonacci sequence (0, 1, 1, 2, 3, 5, 8, 13, 21,...). 
Wite a function that returns the n-th element in the sequence
```{r}
fibo <- function(n){
  x1 <- 0
  x2 <- 1
  if (n == 1) return(x1)
  if (n == 2) return(x2)
  for (i in 3:n) {
    x3 <- x1 + x2
    x1 <- x2
    x2 <- x3
    }
  return(x3)
}
system.time(o1 <- fibo(10000))

# more R-style solution using a recursive function 
fi <- function(n){
  if (n == 1) return(0)
  if (n == 2) return(1)
  fi(n-1) + fi(n-2)
}

# however, the latter function is terribly slow - compare execution time!!
system.time(o1 <- fibo(30))
system.time(o2 <- fi(30))
```

<br/>


#### lapply() and sapply() are for lists and vectors
These powerful functions, along with their close relatives vapply() and tapply(), among others, offer a concise and convenient means of implementing the Split-Apply-Combine strategy for data analysis.

The lapply() function (l stands for list) takes a list as input, applies a function to each element of the list, then returns a list of the same length as the original one. Since a data frame is just a list of vectors (you can see this with as.list(data)), we can use lapply() to apply the class() function to each column of the dataset. 
```{r}
cls_list <- lapply(data, class)
cls_vect <- as.character(cls_list) # chacter vector

cls_vect <- sapply(data, class) # sapply (s stands for simplifying the output) allows to get the above in one command

#calculate mean for each column with numbers
lapply(data[,2:5], mean, na.rm = TRUE) # NA needs to be omitted
sapply(data[,2:5], mean, na.rm = TRUE) # to get a vector
sapply(data[,2:5], mean, na.rm = TRUE, simplify = F) # to get a list as in lapply

#However if lapply returns list of vectors, e.g.
lapply(data[,2:5], range, na.rm = TRUE)
# then sapply returns the matrix
sapply(data[,2:5], range, na.rm = TRUE)

# what if lapply returns vectors of different length?
unique_vals <- lapply(data, unique, na.rm = TRUE)
sapply(unique_vals, length)
# then sapply returns the list, same as lapply
sapply(data, unique, na.rm = TRUE)

# using lapply() with an anonymous function, e.g. to get the first unique elements
lapply(unique_vals, function(x) x[1])

```
<br/>


#### vapply() returns a vector explicitly, while sapply() tries to guess the output format
```{r}
vapply(data, class, character(1)) # this might be a safe option to use in a running script rather than playing with data interactively
```
<br/>


#### tapply() splits data up into groups based on the value of the 2nd listed variable, then applies a function to the members of each group
```{r}
tapply(data$weight, data$sex, summary) # summary of weights by sex
```

<br/>


#### apply() is for matrices and data frames
```{r, results="hide"}
apply(data[,2:5], 2, mean, na.rm = TRUE) # returns a vector of means by columns
apply(data[,2:5], 1, mean, na.rm = TRUE) # returns a vector of means by rows
df <- data[apply(data[,2:5], 1, sum, na.rm = TRUE) > 500,] # select rows with sum above 500

apply(data, 2, function(x) length(unique(na.omit(x)))) # count unique values in each column, omitting NA

apply(data[,2:5], 1, function(x) length(x[x < 80])) # how many values below 80 each row has 
apply(data, 2, function(x) class(x))
```
<br/>

Exploring data
-------------------------
```{r}
object.size(data) # size of a dataset

summary(data)
str(data)
table(data$sex)

```

<br/>


Sampling
------------------
```{r}
sample(1:6, 4, replace = TRUE) # simulate rolling four six-sided dice
sample(LETTERS, 5) # w/o replacement

flips <- sample(c(0, 1), 100, replace = TRUE, prob = c(0.3, 0.7)) #flipping unfair coin 100 times
sum(flips)
flips2 <- rbinom(n = 100, size = 1, prob = 0.7) # same as above
sum(flips2)

rnorm(10, mean = 100, sd = 25)

my_pois <- replicate(100, rpois(5, 10)) # outputs a matrix, each column of which contains 5 random numbers generated from a Poisson distribution with lambda (or mean) of 10
cm <- colMeans(my_pois)
hist(cm) # Central limit theoreme at work!
```
<br/>



Debugging in R
-------------------
* traceback()
* Use browser() to single-step through your code. Place it within your function at the point you want to examine (e.g.) local variables.
* Use debug(function.name) to step through entire function. undebug() will remove that debug call.
* recover() is like browser, except you can choose which level to inspect, rather than the level at which browser was called.
* Add warnings and errors to your code using warning(), stop().
* Can add “assertions” into your code to check that certain values hold; e.g.,
stopifnot (x>0)
* debug(fn), undebug(fn)
* To run a script with all commands and output, in Console of RStudio type source("test.R", echo=T)
<br/>



Running R scripts in command line
-----------------
* At the command line, type “R CMD BATCH trig.R”. R will start up,
process your commands and then quit. Output will be in the file trig.Rout. If there were no errors, the last line of the output file shows the time taken to run the script.
* At the command line, "Rscript trig.R"" will print the output to the terminal
* To run the script as "./trig.R" (for example from the bash script), add "#!/usr/bin/env Rscript" at the top of the file and make "chmod 755 trig.R"
* Another way to run a script, "R < trig.R --no-save", or --save
* Any R command can be run in the command line as 'Rscript -e "getwd()"' or 'Rscript -e "install.pacjages(c("pack1", "pack2"))' , or preceed with "sudo" if it is installed not in the default location
* See Rscript --help for options. For example, --verbose can be used.

<br/>


Using GitHub from R studio
----------------------------
http://happygitwithr.com/new-github-first.html provides instruction how to setup R studio to use with git and Github.

#### “GitHub first, then RStudio” sequence for each new project

* In github browser, create a new repo, with README.
* In R Studio, start a new RStudio Project, via File > New Project > Version Control > Git, opening a new session and creating a new folder on your computer.
* Using Git tab on the right panel, do commit of all valuable changes in local git repository.
* Once ready, push local changes to GitHub, but first, to synchronize with local git repo, pull the current version from GitHub, pushing the blue button "Pull", and then the green button "Push".

<br/>

Using R markdown in github
--------
After http://happygitwithr.com/rmd-test-drive.html
The magical process that turns your R Markdown to HTML is like so: foo.Rmd --> foo.md --> foo.html. Note the intermediate markdown, foo.md. By default RStudio discards this, but you might want to hold on to that markdown.

Why? GitHub gives very special treatment to markdown files. They are rendered in an almost HTML-like way. This is great because it preserves all the charms of plain text but gives you a pseudo-webpage for free when you visit the file in the browser. In contrast, HTML is rendered as plain text on GitHub and you’ll have to take special measures to see it the way you want.

Add in output at the top of Rmd file "keep_md: true"

By making foo.Rmd available, others can see and run your actual code. By sharing foo.md and/or foo.html, others can casually browse your end product and decide if they even want to bother.

* Add figures on GitHub md-document, using for example, ```{r, CLT_50_samples, fig.path='Figures/'} and commiting the folder Figures as well (that is the whole folder with the project)
```{r, CLT_50_samples, fig.path='Figures/'}
n_sample <- 50
lambda <- 8
size <- 5
my_pois <- replicate(n_sample, rpois(size, lambda)) # outputs a matrix, each column of which contains 5 random numbers generated from a Poisson distribution with lambda (or mean) of 10
cm <- colMeans(my_pois)
title <- paste("Central limit theorem at work: \nDistribution of means of ",  n_sample, " samples of size ", size, "\n from the Poisson distribution with lambda = ", lambda)
hist(cm, col = "gainsboro", ylim = c(0, 0.5), xlim = c(2,14), main = title, freq = F, breaks = 20)
lines(density(cm), lwd = 2)
curve(dnorm(x, mean = mean(cm), sd = sd(cm)), col = "red", lwd = 2, add = T)

```

<br/>

```{r, CLT_100_samples, fig.path='Figures/', echo = FALSE}
n_sample <- 100
lambda <- 8
size <- 5
my_pois <- replicate(n_sample, rpois(size, lambda)) # outputs a matrix, each column of which contains 5 random numbers generated from a Poisson distribution with lambda (or mean) of 10
cm <- colMeans(my_pois)
title <- paste("Means of ",  n_sample, " samples")
hist(cm, col = "gainsboro", ylim = c(0, 0.5), xlim = c(2,14), main = title, freq = F, breaks = 20)
lines(density(cm), lwd = 2)
curve(dnorm(x, mean = mean(cm), sd = sd(cm)), col = "red", lwd = 2, add = T)

```

```{r, CLT_1000_samples, fig.path='Figures/', echo = FALSE}
n_sample <- 1000
lambda <- 8
size <- 5
my_pois <- replicate(n_sample, rpois(size, lambda)) # outputs a matrix, each column of which contains 5 random numbers generated from a Poisson distribution with lambda (or mean) of 10
cm <- colMeans(my_pois)
title <- paste("Means of ",  n_sample, " samples")
hist(cm, col = "gainsboro", ylim = c(0, 0.5), xlim = c(2,14), main = title, freq = F, breaks = 20)
lines(density(cm), lwd = 2)
curve(dnorm(x, mean = mean(cm), sd = sd(cm)), col = "red", lwd = 2, add = T)

```

```{r, CLT_10000_samples, fig.path='Figures/', echo = FALSE}
n_sample <- 10000
lambda <- 8
size <- 5
my_pois <- replicate(n_sample, rpois(size, lambda)) # outputs a matrix, each column of which contains 5 random numbers generated from a Poisson distribution with lambda (or mean) of 10
cm <- colMeans(my_pois)
title <- paste("Means of ",  n_sample, " samples")
hist(cm, col = "gainsboro", ylim = c(0, 0.5), xlim = c(4,12), main = title, freq = F, breaks = 20)
lines(density(cm), lwd = 2)
curve(dnorm(x, mean = mean(cm), sd = sd(cm)), col = "red", lwd = 2, add = T)

```

<br/>

Rmd files
-----------------------
http://www.stat.cmu.edu/~cshalizi/rmarkdown/
http://rmarkdown.rstudio.com/html_document_format.html#overview
<br/>

* Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

* In-line code is used like this: In this tutorial, we are using the dataset with `r nrow(data)` data point and `r ncol(data)` variables.

* Name the chunk, like ```{r chunk_name}. This is useful for looking for errors and for figures that will have names based on the name of the chunk

* To pull all of the R code from the Rmd file, use purl("test.Rmd")

* Extract R code and include documentation, use purl("test.Rmd", output = "test.R", documentation = 2)

* To echo the code (but don't run it) from an external script called myscript.R in the Rmarkdown, use {r, code = readLines("myscript.R")} 

* Not to show the code, but show the results, use echo = FALSE.

* Not to show the results, but show the code and the figures, use results = "hide".

* Evaluate the chunk but not to show the code and the result, use include = FALSE.

* To hide figure, use fig.show="hide".

* include=FALSE, all of the code, results, and figures will be suppressed.

* warning=FALSE and message=FALSE suppress any R warnings or messages from being included in the final document.

* fig.path='Figs/' makes it so the figure files get placed in the Figs subdirectory. (By default, they are not saved at all.)

* When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).
