---
title: "R Tips & Tricks"
author: "Julia Ponomarenko, CRG Bioinformatics core facility."
date: "Dec 27, 2017"
output:
  html_document:
    toc: true
    toc_depth: 4
    theme: readable
---

Nice R tutorial
http://journals.plos.org/ploscompbiol/article/file?id=10.1371/journal.pcbi.1000482.s001&type=supplementary
Also swirl tutorials https://github.com/swirldev/swirl_courses. Basic one was used here.

The example dataset used in this tutorial (Davis from the package car)
------------------
```{r}
install.packages("car")
library(car)
data <- Davis
```



Running R scripts in command line
-----------------
* At the command line, type “R CMD BATCH trig.R”. R will start up,
process your commands and then quit.
* Output is stored in the file trig.Rout
* If there were no errors, the last line of the output file shows the time taken to run the script.
* to run a script, use source(’ trig .R’, echo=T) to see commands and output


Files and Directories
--------------------------
```{r}
getwd() # current directory - This is directory where this file is located
# to set up this directory as a working in console
dir <- getwd()
setwd(dir) # run it in console, not from inside the notebook

file.info("R_Tips.Rmd")
file.info("R_Tips.Rmd")$ctime 
list.files()

dir.create("testdir", showWarnings = F)
setwd("testdir") # run in console
file.create("mytest.R")
file.exists("mytest.R")
file.rename("mytest.R","mytest2.R")
file.copy("mytest2.R","mytest3.R")
list.files()

# get full paths of all files and directories containing "R" or "r" in the directory dir
filenames <- list.files(dir, pattern = "*R*", full.names = TRUE, recursive = TRUE, ignore.case = TRUE, include.dirs = TRUE) 
Dir_separator <- "/" # for Windows it is "\"
file_name <- as.character(lapply(strsplit(filenames[1], Dir_separator), tail, n = 1)) # to get just the file name, for example
# while full.names = FALSE in list.files()does the same without the need to parse filenames

list.dirs() # also useful command when looping via directories is needed


dir.create(file.path('testdir2', 'testdir3'), recursive = TRUE)
setwd("..") # one directory up
unlink("testdir", recursive = T)  #removes testdir and all folders inside
```


Read files
-----------
```{r}
file <- "R_Tips.Rmd"
x <- read.table(file, skip = 7, nrows = 1) # for example, we read only line 8
x <- as.numeric(x[2]) # for example, we need a value of the 2nd element of the vector x

# let's write data to the file with the description lines on top
file <- "Davis"
conn <- file(file, open = "w")
writeLines("This is the dataset Davis from the package car", conn)
write("This is my favorite dataset", conn, append = TRUE)
close(conn)

conn <- file(file, open = "a")
write.table(data, conn, row.names = FALSE) # can be write.csv()
close(conn)

#to read a data frame 
conn <- file(file, open = "r")
df <- read.table(conn, skip = 2, header = TRUE) # can be read.csv()
close(conn)


```


Sequences, Numbers, Vectors, Matrices, Lists  
-----------
```{r}
w <<- 10 # assign w to a global environment ???

6.03 / 2
6.03 %/% 2 # integer division
6.03 %% 2 # getting remainder

my_seq <- seq(5,10,length=30)
my_seq

# all three below are identical
1:length(my_seq)
seq(along.with = my_seq)
seq_along(my_seq)

# notice difference!
rep(c(1,2,3), times = 10)
rep(c(1,2,3), each = 10)

x <- c(1,2,3,4,5,6)
y <- x > 3 # boolean vector

chars <- c("my", "name", "is")
x <- paste(chars, collapse = " ") # concatenate vector
paste("my", "name", "is", sep = " ") # same as above, but not paste(chars, sep = " ")

# Careful: paste with vectors recycles the shorter vector
paste(LETTERS, 1:4, sep = "-") # output is a vector!

# random sampling from vectors without replacement
y <- rnorm(1000)
z <- rep(NA, 1000)
my_data <- sample(c(y, z), 100)
my_na <- is.na(my_data)
sum(my_na) # how many NAs
x <- my_data
y <- x[!is.na(x)]
y[y > 0]
x[!is.na(x) & x > 0] # same as above
x[x > 0] # gives NA for NA because NA > 0 = NA

# NaN stand for "not a number"
# Inf stand for infinity

my_vector <- 1:20
dim(my_vector) <- c(4, 5) # dim allows both to get and set the 'dim' attribute to make a vector to a matrix
attributes(my_vector)
class(my_vector)
my_matrix <- my_vector

my_matrix2 <- matrix(data = 1:20, nrow = 4)
identical(my_matrix, my_matrix2)

# naming matrix rows and columns
patients <- c("Bill", "Gina", "Kelly", "Sean") # are in rows
cbind(patients,my_matrix) # numbers in my_matrix are implicitly coerced in characters!
my_data <- data.frame(patients, my_matrix) # so, use a data.frame instead
cnames <- c("patient", "age", "weight", "bp", "rating", "test")
colnames(my_data) <- cnames

# which() function finds the indices:
ints <- sample(10)
which(ints > 7)
any(ints < 0)
all(ints > 0)
```

#### Lists
```{r}
l <- list(id = "Joe", age = 25, dob = c(1980, 3, 29))
l$id
l[[1]]
l["id"] <- "Jim" #change element
l["age"] <- NULL # drop element
```

#### Operators &, &&, | and ||
```{r}
#the `&` operator to evaluate AND across a vector:
TRUE & c(TRUE, FALSE, FALSE)
#The `&&` version of AND only evaluates the first member of a vector:
TRUE && c(TRUE, FALSE, FALSE)

#The OR operator follows a similar set of rules. The `|` version of OR evaluates OR across an entire vector, while the `||` version of OR only evaluates the first member of a vector.
TRUE | c(TRUE, FALSE, FALSE)

#All AND operators are evaluated before OR operators

isTRUE(6 > 4)
xor(5 == 6, !FALSE)

```

Functions
-----------
Type the function name to see its source code.
Use args(function_name) to see the function arguments

#### Function can be passed as an argument to other functions
```{r}
evaluate <- function(func, dat){
  return(func(dat))
}
evaluate(sd, c(1.4, 3.6, 7.9, 8.8))
evaluate(function(x){x+1}, 6)
evaluate(function(x){x[1]}, c(8, 4, 0))
evaluate(function(x){x[length(x)]}, c(8, 4, 0)) # accessing the last element
```

#### Ellipses
Use ellipsis when the function calls the variable number of arguments, see https://www.r-bloggers.com/r-three-dots-ellipsis/

The ellipses can be used to pass on arguments to other functions that are
used within the function you're writing. Usually a function that has the
ellipses as an argument has the ellipses as the last argument. The usage of
such a function would look like:
ellipses_func(arg1, arg2 = TRUE, ...)
Interestingly the usage for the paste function is as follows:
paste (..., sep = " ", collapse = NULL)
Notice that the ellipses is the first argument, and all other arguments after the ellipses have default values. This is a strict rule in R programming: all arguments after an ellipses must have default values. 
```{r}
telegram <- function(...){
  args <- list(...)
  txt <- paste(args, collapse = " ")
  paste("START", txt, "STOP", collapse = " ")
}
telegram("I", "want", "to eat")


telegram2 <- function(...){
  args <- list(...)
  #extract named arguments from the args list by the name of the argument and double brackets: 
  who <- args[["who"]]
  verb <- args[["verb"]]
  what <- args[["what"]]
  paste("The person", who, "decided that she", verb, "to do the following", what)
}
telegram2(who = "I", verb = "want", what = "to eat")

```


Vectorization - Looping through an entire vector
--------------------------
```{r}
x <- runif(100) # a vector of uniformly distributed 100 values
y <- ifelse(x > 0.5, 1, 0) 
```

Q: Flip a biased coin [p=0.6 of heads] 100 times; how many heads do you get? Repeat this for 1000 trials.
```{r}
x <- replicate(1000, rbinom(n = 100, size = 1, p = 0.6))
heads <- colSums(x)
hist(heads)

# or the same
hist(rbinom(n = 1000, size = 100, p = 0.6))
```


Q. Fibonacci sequence (0, 1, 1, 2, 3, 5, 8, 13, 21,...). 
Wite a function that returns n-th element in the sequence
```{r}
fibo <- function(n){
  x1 <- 0
  x2 <- 1
  if (n == 1) return(x1)
  if (n == 2) return(x2)
  for (i in 3:n) {
    x3 <- x1 + x2
    x1 <- x2
    x2 <- x3
    }
  return(x3)
}
system.time(o1 <- fibo(10000))

# more R-style solution using a recursive function 
fi <- function(n){
  if (n == 1) return(0)
  if (n == 2) return(1)
  fi(n-1) + fi(n-2)
}

# how the latter function is terribly slow - compare execution time!!
system.time(o1 <- fibo(30))
system.time(o2 <- fi(30))
```


#### lapply() and sapply() are for lists and vectors
These powerful functions, along with their close relatives (vapply() and tapply(), among others) offer a concise and convenient means of implementing the Split-Apply-Combine strategy for data analysis.

The lapply() function (l stands for list) takes a list as input, applies a function to each element of the list, then returns a list of the same length as the original one. Since a data frame is just a list of vectors (you can see this with as.list(data)), we can use lapply() to apply the class() function to each column of the dataset. 
```{r}
cls_list <- lapply(data, class)
cls_vect <- as.character(cls_list) # chacter vector

cls_vect <- sapply(data, class) # sapply (s stands for simplifying the output) allows to get the above in one command

#calculate mean for each column with numbers
lapply(data[,2:5], mean, na.rm = TRUE) # NA needs to be omitted
sapply(data[,2:5], mean, na.rm = TRUE) # to get a vector
sapply(data[,2:5], mean, na.rm = TRUE, simplify = F) # to get a list as in lapply

#However if lapply returns list of vectors, e.g.
lapply(data[,2:5], range, na.rm = TRUE)
# then sapply returns the matrix
sapply(data[,2:5], range, na.rm = TRUE)

# what if lapply returns vectors of different length?
unique_vals <- lapply(data, unique, na.rm = TRUE)
sapply(unique_vals, length)
# then sapply returns the list, same as lapply
sapply(data, unique, na.rm = TRUE)

# using lapply() with an anonymous function, e.g. to get the first unique elements
lapply(unique_vals, function(x) x[1])

```

#### vapply() returns a vector explicitly, while sapply() tries to guess the output format
```{r}
vapply(data, class, character(1)) # this might be a safe option to use in a running script rather than playing with data interactively
```

#### tapply() splits data up into groups based on the value of some variable, then applies a function to the members of each group.
```{r}
tapply(data$weight, data$sex, summary) # summary of weights by sex
```


#### apply() is for matrices and data frames
```{r}
apply(data[,2:5], 2, mean, na.rm = TRUE) # returns a vector of means by columns
apply(data[,2:5], 1, mean, na.rm = TRUE) # returns a vector of means by rows
df <- data[apply(data[,2:5], 1, sum, na.rm = TRUE) > 500,] # select rows with sum above 500

apply(data, 2, function(x) length(unique(na.omit(x)))) # count unique values in each column, omitting NA

apply(data[,2:5], 1, function(x) length(x[x < 80])) # how many values below 80 each row has 
apply(data, 2, function(x) class(x))
```

Exploring data
-------------------------
```{r}
object.size(data) # size of a dataset

summary(data)
str(data)
table(data$sex)

```


Sampling
------------------
```{r}
sample(1:6, 4, replace = TRUE) # simulate rolling four six-sided dice
sample(LETTERS, 5) # w/o replacement

flips <- sample(c(0, 1), 100, replace = TRUE, prob = c(0.3, 0.7)) #flipping unfair coin 100 times
sum(flips)
flips2 <- rbinom(n = 100, size = 1, prob = 0.7) # same as above
sum(flips2)

rnorm(10, mean = 100, sd = 25)

my_pois <- replicate(100, rpois(5, 10)) # outputs a matrix, each column of which contains 5 random numbers generated from a Poisson distribution with lambda (or mean) of 10
cm <- colMeans(my_pois)
hist(cm) # Central limit theoreme at work!
```


Debugging in R
-------------------
* traceback()
* Use browser() to single-step through your code. Place it within your function at the point you want to examine (e.g.) local variables.
* Use debug(function.name) to step through entire function. undebug() will remove that debug call.
* recover() is like browser, except you can choose which level to inspect, rather than the level at which browser was called.
* Add warnings and errors to your code using warning(), stop().
* Can add “assertions” into your code to check that certain values hold; e.g.,
stopifnot (x>0)
* debug(fn), undebug(fn)

Bioconductor
--------------------

Access to databases
--------------------

Paralllelization
--------------------

Using GitHub from R studio
----------------------------
http://happygitwithr.com/new-github-first.html provides instruction how to setup R studio to use with git and Github.
#### Follow this “GitHub first, then RStudio” sequence for each new project:
## In github browser, create a new repo with README
## In R Studio, start a new RStudio Project, via File > New Project > Version Control > Git, opening a new session and creating a new folder on your computer.
## Using Git tab on the right panel, do commit of all valuable changes in local git repository.
## Push local changes to GitHub, but first pull the current version from GitHub, pushing the blue button "Pull"




Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).
